# Alpha diversity

To investigate the effect of fragmentation on leaf-litter ant alpha diversity per sampling point, we fitted a generalized linear mixed-effects model with the number of patches in the landscape as the predictor and island and landscape as random effects.

## Loading packages and importing data

```{r, output=FALSE}
#| code-overflow: wrap
library(dplyr)
library(ggplot2)
library(ggimage)
library(MuMIn)
library(performance)
library(lmerTest)


# Importing data
rm(list = ls()); gc()
set.seed(13)

comm <- read.csv("ants_comm.csv", header = T)
env <- read.csv("sampled_lands.csv", header = T)

complete <- merge(comm, env, by = "point_ID")
complete <- complete[,c(1,147:153,2:145)]
```

## Calculating species richness by sampling point (alpha diversity)

```{r, output=FALSE}
#| code-overflow: wrap
complete[,c(9:152)] <- 
  complete[,c(9:152)] %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))

complete$richness <- rowSums(complete[,9:152]) # richness by point
```

## Fitting the model

```{r}
#| code-overflow: wrap
mod1 <- glmer(richness ~ number_patches + (1 + number_patches | landscape),
              data = complete, family = "poisson")
summary(mod1) #slope=-0.01, p < 0.001
check_residuals(mod1)
check_overdispersion(mod1)
```

## Plotting the graph

```{r}
#| code-overflow: wrap

# Identifying the level of fragmentation in the landscapes
complete$cor <- NA
complete$cor[1:80] <- "#4575b4"
complete$cor[81:120] <- "#fee090"
complete$cor[121:200] <- "#d73027"


# Calculating the predicted values to plot the graph
newdata <- data.frame(number_patches = seq(min(complete$number_patches),
                       max(complete$number_patches),
                       length.out = 100))

newdata$pred <- predict(mod1, newdata = newdata,
                        type = "response",
                        re.form = NA)

pred_link <- predict(mod1, newdata = newdata,
                     type = "link",
                     se.fit = TRUE,
                     re.form = NA)

newdata$fit <- exp(pred_link$fit)
newdata$lwr <- exp(pred_link$fit - 1.96 * pred_link$se.fit)
newdata$upr <- exp(pred_link$fit + 1.96 * pred_link$se.fit)

plot_alpha <- 
ggplot(complete, aes(x = number_patches, y = richness)) +
  
  geom_point(size = 4, shape = 21,
             color = "#252525",
             fill = complete$cor,
             alpha = 0.6) +
  
  geom_ribbon(data = newdata,
              aes(x = number_patches,
                  ymin = lwr,
                  ymax = upr),
              inherit.aes = FALSE,
              fill = "#cccccc",
              alpha = 0.5) +
  
  geom_line(data = newdata,
            aes(x = number_patches, y = fit),
            inherit.aes = FALSE,
            color = "#252525",
            linewidth = 1) +
  
  labs(x = "Number of islands in the landscape",
       y = "Number of ant species in the sampling points") +
  
  geom_image(
    data = tibble(number_patches = 55, richness = 19.2),
    aes(image = "ant_1.png"), size = 0.08 
  ) + # https://www.phylopic.org/images/3ae7cb42-8904-4775-87d6-968255913711/camponotus-vagus
  
  theme_bw(base_size = 10) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "#252525"),
        axis.title = element_text(colour = "#252525", face = "bold", size = 14),
        axis.title.x = element_text(vjust = 0, size = 14),
        axis.title.y = element_text(vjust = 2, size = 14),
        axis.text = element_text(colour = "#252525"),
        axis.ticks = element_line(colour = "#252525", size = 0.25))

plot_alpha

ggsave(
  plot = plot_alpha,
  filename = "alpha_diversity_NumberPatches.png", dpi = 500,
  width = 8 * 1.6, height = 8 * 1.7, units = 'cm') 
```
