# Limited dispersal among patches

To confirm if there was low movement of individuals among patches in our system, we compared the maximum flight distance of the winged ant queens of each species in our dataset with the mean nearest-neighbour distances between islands in our landscapes. 

We inferred the maximum flight distance of the ant queens using a linear regression between the head width of winged ant queens and their maximum flight distance estimated by [Helms 2018](https://jacksonhelms.com/wp-content/uploads/2019/04/helms-2018-ant-flight-ecology-1.pdf): maximum flight distance = 2.03 * log10(winged queen head width) + 2.37. We used head width measurements of winged queens available on [AntWiki](https://www.antwiki.org/wiki/) and calculated the community-weighted mean of the inferred maximum flight distance for each landscape. We then calculated the mean nearest-neighbour distance between sampled islands and their nearest islands in each landscape to compare between-patch distances to the maximum flight distance estimates.

## Flight capacity of winged ant queens
### Loading packages and importing data
```{r, output=FALSE}
library(BAT)
library(dplyr)

# Importing data
rm(list = ls()); gc()
set.seed(13)

comm <- read.csv("ants_comm.csv", header = T)
env <- read.csv("sampled_lands.csv", header = T)

complete <- merge(comm, env, by = "point_ID")
complete <- complete[,c(1,146:154,2:145)]

comm_landscape <- complete[,-c(1,2,6)] %>%
  group_by(landscape) %>%
  summarise(
    across(c(number_patches, forest_cover, long, lat), first),
    across(c(patch_area), mean), #mean patch size by landscape
    across(
      where(is.numeric) & 
        !all_of(c("number_patches", "forest_cover", "long", "lat", "patch_area")),
      sum
    ),
    .groups = "drop"
  )

comm_landscape <- comm_landscape[, -c(2:6)]
remove(comm)
remove(complete)
remove(env)

comm_landscape <- as.data.frame(comm_landscape)
rownames(comm_landscape) <- comm_landscape$landscape
comm_landscape <- comm_landscape[,-1]

traits <- read.csv("ant_traits.csv")
setdiff(traits$especie, colnames(comm_landscape)[1:144])
traits$winged_queen <- as.factor(traits$winged_queen)

winged_queens <- subset(traits, winged_queen == "Yes")
remove(traits)
```

I used the regression equation from Helms (2018) to estimate the flight distance of winged ant species and then calculated the community-weighted mean (CWM) estimated flight distance for each landscape.

### Estimating flight distance of winged ant species 
```{r, output=FALSE}
winged_queens$max_flight_dist <- 2.03 *log10(winged_queens$queen_HeadWidth) + 2.37
winged_queens$max_flight_dist <- 10^winged_queens$max_flight_dist

winged_queens <- winged_queens[,-c(2:7)]
winged_queens$especie <- gsub(" ", "_", winged_queens$especie)

extra_cols <- setdiff(colnames(comm_landscape), winged_queens$especie) # Identify which species should be removed from the community column. Note that I selected only species with winged queens.
comm_landscape <- comm_landscape[, !(names(comm_landscape) %in% extra_cols)]

rownames(winged_queens) <- winged_queens$especie
winged_queens <- as.data.frame(winged_queens[-1])
```


### Calculate CWM of maximum flight distance
```{r}
#| code-overflow: wrap
CWM_max_flight_dist <- cwm(comm = comm_landscape, trait = winged_queens, abund = TRUE, na.rm = TRUE)
CWM_max_flight_dist <- as.data.frame(CWM_max_flight_dist)
rownames(CWM_max_flight_dist) <- rownames(comm_landscape)

comm_landscape$landscape <- row.names(comm_landscape)
CWM_max_flight_dist$landscape <- row.names(CWM_max_flight_dist)

comm_landscape <- merge(comm_landscape, CWM_max_flight_dist, by = "landscape")
comm_landscape <- comm_landscape[,c(1,94,2:93)]

comm_landscape[,c(3:94)] <- 
  comm_landscape[,c(3:94)] %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))

comm_landscape$richness <- rowSums(comm_landscape[,-c(1,2)]) # richness by landscape

comm_landscape$type <- NA
comm_landscape$type[1:8] <- "FL"
comm_landscape$type[9:12] <- "MS"
comm_landscape$type[13:20] <- "SS"
comm_landscape$type <- as.factor(comm_landscape$type)

comm_landscape <- comm_landscape[,c(1,2,95,96, 3:94)]
comm_landscape$max_flight_dist <- round(comm_landscape$max_flight_dist, 2)

comm_landscape
```

## Mean Nearest Neighbour Distance
### Loading packages and importing data
```{r, output=FALSE}
#|code-overflow: wrap
library(raster)
library(sf)
library(terra)
library(landscapemetrics)
library(tidyr)

Balbina <- raster("Balbina_mapbiomas_collection8_2022.tif")
new_crs <- CRS("+proj=utm +south +zone=21 +datum=WGS84 +units=m +no_defs")
Balbina <- projectRaster(Balbina, crs=new_crs)
values(Balbina)  = as.integer(values(Balbina))
check_landscape(Balbina)

# Select only forest pixels 
forest_formation <- Balbina == 3
forest_formation <- projectRaster(forest_formation, crs=new_crs)
values(forest_formation)  = as.integer(values(forest_formation))
check_landscape(forest_formation)
remove(Balbina)

mean_dist_landscapes <- data.frame()
```

### Calculating the mean distance between the sampled patches x all patches in the landscape
```{r, output=FALSE}
#| code-overflow: wrap
#### Small landscapes ####
setwd("C:/Users/ivana/OneDrive/PhD_INPA/2.SLOSS_question/Code_paper_SLOSS/individual_landscapes")

S1 <- st_read("S1.shp")
S1 <- st_transform(S1, crs = new_crs)
S2 <- st_read("S2.shp")
S2 <- st_transform(S2, crs = new_crs)
S3 <- st_read("S3.shp")
S3 <- st_transform(S3, crs = new_crs)
S4 <- st_read("S4.shp")
S4 <- st_transform(S4, crs = new_crs)
S7 <- st_read("S7.shp")
S7 <- st_transform(S7, crs = new_crs)
S8 <- st_read("S8.shp")
S8 <- st_transform(S8, crs = new_crs)
S9 <- st_read("S9.shp")
S9 <- st_transform(S9, crs = new_crs)
S10 <- st_read("S10.shp")
S10 <- st_transform(S10, crs = new_crs)

setwd("C:/Users/ivana/OneDrive/PhD_INPA/2.SLOSS_question/Code_paper_SLOSS")
point_ants <- read.csv("ants_env.csv")
point_ants <- st_as_sf(point_ants, coords = c("long_utm", "lat_utm"), crs = new_crs)

# S1
S1_raster <- crop(forest_formation, S1)
S1_raster <- mask(S1_raster, S1)
S1_raster <- rast(S1_raster)

patches <- get_patches(S1_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S1_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")
media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- data.frame(
  id = "S1",
  mean_dist = media_paisagem
)

# S2
S2_raster <- crop(forest_formation, S2)
S2_raster <- mask(S2_raster, S2)
S2_raster <- rast(S2_raster)

patches <- get_patches(S2_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S2_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S2",
  mean_dist = media_paisagem
)

# S3
S3_raster <- crop(forest_formation, S3)
S3_raster <- mask(S3_raster, S3)
S3_raster <- rast(S3_raster)

patches <- get_patches(S3_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S3_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S3",
  mean_dist = media_paisagem
)

# S4
S4_raster <- crop(forest_formation, S4)
S4_raster <- mask(S4_raster, S4)
S4_raster <- rast(S4_raster)

patches <- get_patches(S4_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S4_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S4",
  mean_dist = media_paisagem
)

# S7
S7_raster <- crop(forest_formation, S7)
S7_raster <- mask(S7_raster, S7)
S7_raster <- rast(S7_raster)

patches <- get_patches(S7_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S7_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S7",
  mean_dist = media_paisagem
)

# S8
S8_raster <- crop(forest_formation, S8)
S8_raster <- mask(S8_raster, S8)
S8_raster <- rast(S8_raster)

patches <- get_patches(S8_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S8_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S8",
  mean_dist = media_paisagem
)

# S9
S9_raster <- crop(forest_formation, S9)
S9_raster <- mask(S9_raster, S9)
S9_raster <- rast(S9_raster)

patches <- get_patches(S9_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S9_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S9",
  mean_dist = media_paisagem
)

# S10
S10_raster <- crop(forest_formation, S10)
S10_raster <- mask(S10_raster, S10)
S10_raster <- rast(S10_raster)

patches <- get_patches(S10_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(S10_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "S10",
  mean_dist = media_paisagem
)

#### Medium-sized landscapes ####
setwd("C:/Users/ivana/OneDrive/PhD_INPA/2.SLOSS_question/Code_paper_SLOSS/individual_landscapes")

M2 <- st_read("M2.shp")
M2 <- st_transform(M2, crs = new_crs)
M6 <- st_read("M6.shp")
M6 <- st_transform(M6, crs = new_crs)
M9 <- st_read("M9.shp")
M9 <- st_transform(M9, crs = new_crs)
M10 <- st_read("M10.shp")
M10 <- st_transform(M10, crs = new_crs)

# M2
M2_raster <- crop(forest_formation, M2)
M2_raster <- mask(M2_raster, M2)
M2_raster <- rast(M2_raster)

patches <- get_patches(M2_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(M2_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "M2",
  mean_dist = media_paisagem
)


# M6
M6_raster <- crop(forest_formation, M6)
M6_raster <- mask(M6_raster, M6)
M6_raster <- rast(M6_raster)

patches <- get_patches(M6_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(M6_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "M6",
  mean_dist = media_paisagem
)


# M9
M9_raster <- crop(forest_formation, M9)
M9_raster <- mask(M9_raster, M9)
M9_raster <- rast(M9_raster)

patches <- get_patches(M9_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(M9_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "M9",
  mean_dist = media_paisagem
)


# M10
M10_raster <- crop(forest_formation, M10)
M10_raster <- mask(M10_raster, M10)
M10_raster <- rast(M10_raster)

patches <- get_patches(M10_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(M10_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "M10",
  mean_dist = media_paisagem
)


#### Few large landscapes ####
L1 <- st_read("L1.shp")
L1 <- st_transform(L1, crs = new_crs)
L2 <- st_read("L2.shp")
L2 <- st_transform(L2, crs = new_crs)
L3 <- st_read("L3.shp")
L3 <- st_transform(L3, crs = new_crs)
L4 <- st_read("L4.shp")
L4 <- st_transform(L4, crs = new_crs)
L7 <- st_read("L7.shp")
L7 <- st_transform(L7, crs = new_crs)
L8 <- st_read("L8.shp")
L8 <- st_transform(L8, crs = new_crs)
L9 <- st_read("L9.shp")
L9 <- st_transform(L9, crs = new_crs)
L10 <- st_read("L10.shp")
L10 <- st_transform(L10, crs = new_crs)


# L1
L1_raster <- crop(forest_formation, L1)
L1_raster <- mask(L1_raster, L1)
L1_raster <- rast(L1_raster)

patches <- get_patches(L1_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L1_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L1",
  mean_dist = media_paisagem
)


# L2
L2_raster <- crop(forest_formation, L2)
L2_raster <- mask(L2_raster, L2)
L2_raster <- rast(L2_raster)

patches <- get_patches(L2_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L2_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L2",
  mean_dist = media_paisagem
)


# L3
L3_raster <- crop(forest_formation, L3)
L3_raster <- mask(L3_raster, L3)
L3_raster <- rast(L3_raster)

patches <- get_patches(L3_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L3_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L3",
  mean_dist = media_paisagem
)


# L4
L4_raster <- crop(forest_formation, L4)
L4_raster <- mask(L4_raster, L4)
L4_raster <- rast(L4_raster)

patches <- get_patches(L4_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L4_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L4",
  mean_dist = media_paisagem
)


# L7
L7_raster <- crop(forest_formation, L7)
L7_raster <- mask(L7_raster, L7)
L7_raster <- rast(L7_raster)

patches <- get_patches(L7_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L7_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L7",
  mean_dist = media_paisagem
)


# L8
L8_raster <- crop(forest_formation, L8)
L8_raster <- mask(L8_raster, L8)
L8_raster <- rast(L8_raster)

patches <- get_patches(L8_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L8_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L8",
  mean_dist = media_paisagem
)


# L9
L9_raster <- crop(forest_formation, L9)
L9_raster <- mask(L9_raster, L9)
L9_raster <- rast(L9_raster)

patches <- get_patches(L9_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L9_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L9",
  mean_dist = media_paisagem
)


# L10
L10_raster <- crop(forest_formation, L10)
L10_raster <- mask(L10_raster, L10)
L10_raster <- rast(L10_raster)

patches <- get_patches(L10_raster, directions = 8, class = 1, return_raster = TRUE)
patches_raster <- patches$layer_1$class_1
point_ants$point_ID2 <- terra::extract(patches_raster, point_ants)[,2]

enn_patches <- lsm_p_enn(L10_raster, directions = 8)
enn_patches <- subset(enn_patches, enn_patches$class == 1)
pontos_enn <- merge(point_ants, enn_patches, by.x = "point_ID2", by.y = "id")

media_paisagem <- mean(pontos_enn$value)
mean_dist_landscapes <- dplyr::add_row(
  mean_dist_landscapes,
  id = "L10",
  mean_dist = media_paisagem
)

colnames(mean_dist_landscapes)[1] <- "landscape"
colnames(mean_dist_landscapes)[2] <- "mean_dist_sampled_patches"

comm_landscape <- merge(comm_landscape, mean_dist_landscapes, by = "landscape")
comm_landscape <- comm_landscape[,c(1,3,2,97)]
```

```{r}
comm_landscape
```