# Changes in landscape size

To evaluate the robustness of our results to changes in landscape size, we conducted two analyses:

1) We enlarged the landscapes by adding 2 km and 4 km around the edges, creating landscapes of 6 x 6 km and 8 x 8 km, and repeated the gamma diversity analysis. 
2) We fit simple linear regressions relating point-scale ant richness (alpha diversity) to habitat amount and fragmentation in landscapes defined as circular buffers surrounding the points. We did this at multiple spatial extents, from 150-m to 2000-m radius buffers in 50 m increments.

## Landscapes of 6x6km and 8x8km
### Loading packages and importing data
```{r, output=FALSE}
#| code-overflow: wrap
rm(list = ls()); gc()
set.seed(13)

# Load packages
library(raster)
library(terra)
library(landscapemetrics)
library(ggplot2)
library(dplyr)
library(MuMIn)
library(ggimage)
library(sf)

setwd("C:/Users/ivana/OneDrive/PhD_INPA/2.SLOSS_question/Code_paper_SLOSS")
Balbina <- raster("Balbina_mapbiomas_collection8_2022.tif")
new_crs <- CRS("+proj=utm +south +zone=21 +datum=WGS84 +units=m +no_defs")
Balbina <- projectRaster(Balbina, crs = new_crs)
values(Balbina) = as.integer(values(Balbina))
check_landscape(Balbina)

forest_formation <- Balbina == 3
forest_formation <- projectRaster(forest_formation, crs = new_crs)
values(forest_formation)  = as.integer(values(forest_formation))
check_landscape(forest_formation)
```

### 6x6 km landscapes: Calculating forest amount and number of islands
```{r, output=FALSE}
#| code-overflow: wrap
landscapes6km <- vect("6km_paisagem.shp")
names(landscapes6km)

forest_cover <- landscapemetrics::sample_lsm(
  forest_formation,
  y = landscapes6km,
  what = "lsm_c_pland",
  directions = 8,
  plot_id = landscapes6km$name
)
forest_cover <- subset(forest_cover, forest_cover$class == 1)  # Select only the presence of forest

number_patches <- landscapemetrics::sample_lsm(forest_formation, 
                                               y = landscapes6km,
                                               what = "lsm_c_np",
                                               directions = 8,
                                               plot_id = landscapes6km$name)

number_patches <- subset(number_patches, number_patches$class == 1)  # Select only the presence of patches

number_patches$plot_id == forest_cover$plot_id
metrics <- as.data.frame(cbind(number_patches$plot_id, number_patches$value,
                               forest_cover$value))
colnames(metrics) <- c("id", "number_patches", "forest_cover")

metrics$type <- NA
metrics$type[1:8] <- "#4575b4" #landscapes with few large patches
metrics$type[9:12] <- "#fee090" #landscapes with several medium-sized patches
metrics$type[13:20] <- "#d73027" #landscapes with many small patches

comm <- read.csv("ants_comm.csv", header = T)
comm <- comm[-201,]
env <- read.csv("sampled_lands.csv", header = T)

complete <- merge(comm, env, by = "point_ID")
complete <- complete[,c(1,147:154,2:145)]

# Calculating gamma diversity - species richness in the landscapes
complete$landscape <- as.factor(complete$landscape)

comm_landscape <- complete[,-c(1,5)] %>%
  group_by(landscape) %>%
  summarise(
    across(c(number_patches, forest_cover, long, lat), first),
    across(c(patch_area), mean), #mean patch size by landscape
    across(
      where(is.numeric) & 
        !all_of(c("number_patches", "forest_cover", "long", "lat", "patch_area")),
      sum
    ),
    .groups = "drop"
  )

comm_landscape[,c(7:150)] <- 
  comm_landscape[,c(7:150)] %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))
```

### 6x6 km landscapes: Fitting the models
```{r}
#|code-overflow: wrap
any(comm_landscape[, 7:150] > 1) #checking if any value is higher than 1 (presence)

comm_landscape$richness <- rowSums(comm_landscape[,7:150]) # richness by landscape

landscapes <- comm_landscape[,c(1:3,6,151)]
colnames(landscapes)[4] <- "mean_patch_area"
comm_landscape$number_patches6km <- as.numeric(metrics$number_patches)
comm_landscape$forest_cover6km <- as.numeric(metrics$forest_cover)

cor.test(comm_landscape$number_patches6km, comm_landscape$forest_cover6km) #cor = -0.28

mod_global6km <- lm(richness ~ number_patches6km + forest_cover6km, data = comm_landscape, na.action = na.fail)
summary(mod_global6km)

models1 <- dredge(mod_global6km)
models1
```

### 6X6 km landscapes: Plotting the graph
```{r}
#|code-overflow: wrap
comm_landscape$cor <- NA
comm_landscape$cor[1:8] <- "#4575b4"
comm_landscape$cor[9:12] <- "#fee090"
comm_landscape$cor[13:20] <- "#d73027"

plot_6km <-
  ggplot(data = comm_landscape,
         mapping = aes(x = number_patches6km, y = richness)) +
  
  geom_smooth(method = "lm", fill = "#cccccc", alpha = 0.5,
              color = "#252525") +
  
  labs(x = "Number of islands in the landscape (6 x 6 km)",
       y = "Number of ant species") +
  
  geom_image(
    data = tibble(number_patches6km = 120, richness = 51),
    aes(image = "ant_1.png"), size = 0.08 
  ) + # https://www.phylopic.org/images/3ae7cb42-8904-4775-87d6-968255913711/camponotus-vagus
  
  geom_point(size = 4, shape = 21,
             color = "#252525", fill = comm_landscape$cor,
             alpha = 0.7) +
  
  theme_bw(base_size = 10) +
  
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "#252525"),
        axis.title = element_text(colour = "#252525", face = "bold", size = 14),
        axis.title.x = element_text(vjust = 0, size = 14),
        axis.title.y = element_text(vjust = 2, size = 14),
        axis.text = element_text(colour = "#252525"),
        axis.ticks = element_line(colour = "#252525", size = 0.25))
plot_6km
```

### 8x8 km landscapes: Calculating forest amount and number of islands
```{r, output=FALSE}
#| code-overflow: wrap
landscapes8km <- vect("paisagem_da_paisagem.shp")
names(landscapes8km)

forest_cover <- landscapemetrics::sample_lsm(
  forest_formation,
  y = landscapes8km,
  what = "lsm_c_pland",
  directions = 8,
  plot_id = landscapes8km$name
)
forest_cover <- subset(forest_cover, forest_cover$class == 1)  # Select only the presence of forest

number_patches <- landscapemetrics::sample_lsm(forest_formation, 
                                               y = landscapes8km,
                                               what = "lsm_c_np",
                                               directions = 8,
                                               plot_id = landscapes8km$name)

number_patches <- subset(number_patches, number_patches$class == 1)  # Select only the presence of patches

number_patches$plot_id == forest_cover$plot_id
metrics2 <- as.data.frame(cbind(number_patches$plot_id, number_patches$value,
                               forest_cover$value))
colnames(metrics2) <- c("id", "number_patches", "forest_cover")

comm_landscape$number_patches8km <- as.numeric(metrics2$number_patches)
comm_landscape$forest_cover8km <- as.numeric(metrics2$forest_cover)
cor.test(comm_landscape$number_patches8km, comm_landscape$forest_cover8km)
```

### 8x8 km landscapes: Fitting the models
```{r}
#|code-overflow: wrap
mod_global8km <- lm(richness ~ number_patches8km + forest_cover8km, data = comm_landscape, na.action = na.fail)
summary(mod_global8km)
models2 <- dredge(mod_global8km)
models2
```

### 8X8 km landscapes: Plotting the graph
```{r}
#|code-overflow: wrap
plot_8km <-
  ggplot(data = comm_landscape,
         mapping = aes(x = number_patches8km, y = richness)) +
  
  geom_smooth(method = "lm", fill = "#cccccc", alpha = 0.5,
              color = "#252525") +
  
  labs(x = "Number of islands in the landscape (8 x 8 km)",
       y = "Number of ant species") +
  
  geom_image(
    data = tibble(number_patches8km = 190, richness = 51),
    aes(image = "ant_1.png"), size = 0.08 
  ) + # https://www.phylopic.org/images/3ae7cb42-8904-4775-87d6-968255913711/camponotus-vagus
  
  geom_point(size = 4, shape = 21,
             color = "#252525", fill = comm_landscape$cor,
             alpha = 0.7) +
  
  theme_bw(base_size = 10) +
  
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "#252525"),
        axis.title = element_text(colour = "#252525", face = "bold", size = 14),
        axis.title.x = element_text(vjust = 0, size = 14),
        axis.title.y = element_text(vjust = 2, size = 14),
        axis.text = element_text(colour = "#252525"),
        axis.ticks = element_line(colour = "#252525", size = 0.25))
plot_8km
```

## Circular buffers surroundind the points
### Loading packages and importing data
```{r, output=FALSE}
#| code-overflow: wrap
data <- read.csv("ants_env.csv")
comm <- read.csv("ants_comm.csv", row.names = 1)
comm <- comm[-201,]
richness <- as.data.frame(rowSums(ifelse(comm > 0, 1, 0)))
richness$point_ID <- row.names(richness)
colnames(richness)[1] <- "richness"

new_crs <- CRS("+proj=utm +south +zone=21 +datum=WGS84 +units=m +no_defs")
point_ants <- st_as_sf(data, coords = c("long_utm", "lat_utm"), crs = new_crs)
point_ants$landscape <- as.factor(point_ants$landscape)

forest <- raster("forest_formation.tif")

any(comm > 1) #checking if any value is higher than 1 (presence)


point_ants$richness <- NA
point_ants$richness <-
  richness$richness[match(point_ants$point_ID, richness$point_ID)]
```

### Calculating forest amount and number of islands inside the buffers
```{r, output=FALSE}
#| code-overflow: wrap
buffers <- seq(150, 2000, by = 50)

for (b in buffers) {
  
  result <- landscapemetrics::sample_lsm(
    forest,
    y = point_ants,
    plot_id = point_ants$point_ID,
    size = b,
    shape = "circle",
    what = "lsm_c_pland",
    directions = 8
  )
  
  # selecting only forest
  result <- subset(result, class == 1)
  result$value <- scale(result$value)

  # colname
  nome_coluna <- paste0("forest_cover_", b)
  
  # adding to the dataset
  point_ants[[nome_coluna]] <- 
    result$value[match(point_ants$point_ID, result$plot_id)]
}


for (b in buffers) {
  result_2 <- landscapemetrics::sample_lsm(
    forest,
    y = point_ants,
    plot_id = point_ants$point_ID,
    size = b,
    shape = "circle",
    what = "lsm_c_np",
    directions = 8
  )
  
  result_2 <- subset(result_2, class == 1)
  result_2$value <- scale(result_2$value)
  
  nome_coluna <- paste0("number_patches_", b)
  
  point_ants[[nome_coluna]] <- 
    result_2$value[match(point_ants$point_ID, result_2$plot_id)]
}
```

### Calculating multiple regressions with forest amount and number of patches on the same scale
```{r}
#|code-overflow: wrap

reg_mult <- lapply(buffers, function(b) {
  
  forest_cover <- paste0("forest_cover_", b)
  number_patch <- paste0("number_patches_", b)
  
  formula <- as.formula(
    paste("richness ~", forest_cover, "+", number_patch)
  )
  
  mod <- lm(formula, data = point_ants)
  s <- summary(mod)
  
  coefs <- as.data.frame(coef(s))
  coefs$term <- rownames(coefs)
  
  coefs$buffer <- b
  coefs$r2 <- s$r.squared
  coefs$r2_adj <- s$adj.r.squared
  coefs$p_model <- pf(s$fstatistic[1], 
                      s$fstatistic[2], 
                      s$fstatistic[3], 
                      lower.tail = FALSE)
  
  rownames(coefs) <- NULL
  coefs
})

reg_mult_df <- bind_rows(reg_mult)

print(reg_mult_df)
```

### Scale of effect: model with higher adjusted r2
```{r}
reg_mult_df[order(-reg_mult_df$r2_adj),][1,]  # 1550 m
```

### Plotting the graph
```{r}
#| code-overflow: wrap

# Selecting only the results for number of islands
reg_mult_df_2 <- reg_mult_df[grepl("^number_patches_", reg_mult_df$term), ]

plot1 <-
  ggplot(reg_mult_df_2 , aes(x = buffer, y = Estimate)) +
  geom_errorbar(aes(ymin = Estimate - `Std. Error`, ymax = Estimate + `Std. Error`), color = 'black', width = 50) +
  geom_point(size = 4) +
  labs(x = "Buffer size (m)",
       y = "Standardized coefficient (SE)",
       title = "Fragmentation (number of  islands)") +
  theme_bw(base_size = 16) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "black"),
        plot.margin = margin(0.5, 1.5, 0.5, 1.5, "cm"))
plot1
```
