# Gamma diversity

To investigate the effect of forest fragmentation on the total number of leaf-litter ant species across the 10 points in each landscape, we performed simple linear regression analysis.

## Loading packages and importing data
```{r, output=FALSE}
#| code-overflow: wrap
library(dplyr)
library(car)
library(ggplot2)
library(tidyverse)
library(ggtext)
library(ggimage)
library(lmtest)
library(grid)
library(png)
library(MuMIn)

# Importing data
rm(list = ls()); gc()
set.seed(13)

comm <- read.csv("ants_comm.csv", header = T)
env <- read.csv("sampled_lands.csv", header = T)

complete <- merge(comm, env, by = "point_ID")
complete <- complete[,c(1,147:154,2:145)]
```

## Calculating the number of species in the landscapes (gamma diversity)
```{r, output=FALSE}
#| code-overflow: wrap
complete$landscape <- as.factor(complete$landscape)

comm_landscape <- complete[,-c(1,5)] %>%
  group_by(landscape) %>%
  summarise(
    across(c(number_patches, forest_cover, long, lat), first),
    across(c(patch_area), mean), #mean patch size by landscape
    across(
      where(is.numeric) & 
        !all_of(c("number_patches", "forest_cover", "long", "lat", "patch_area")),
      sum
    ),
    .groups = "drop"
  )

comm_landscape[,c(7:150)] <- 
  comm_landscape[,c(7:150)] %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))
```

```{r}
any(comm_landscape[, 7:150] > 1) #checking if any value is higher than 1 (presence)
```

```{r, output=FALSE}
#| code-overflow: wrap
comm_landscape$richness <- rowSums(comm_landscape[,7:150]) # richness by landscape

landscapes <- comm_landscape[,c(1:3,6,151)]
colnames(landscapes)[4] <- "mean_patch_area"

landscapes$type <- NA
landscapes$type[1:8] <- "FL"
landscapes$type[9:12] <- "MS"
landscapes$type[13:20] <- "SS"
landscapes$type <- as.factor(landscapes$type)
```

## Testing the correlation between habitat amount and number of patches in the landscapes
One specific condition mentioned by [Fahrig et al. (2022)](https://onlinelibrary.wiley.com/doi/10.1111/brv.12792) is that the total amount of habitat should either be the same across the sampled landscapes, or at least not be correlated with the number of patches. We tested the correlation between forest cover (habitat amount) and number of patches and found no significant correlation:

```{r}
#| code-overflow: wrap
# Testing the correlation between forest cover and number of patches 
# of the sampled landscapes
cor.test(comm_landscape$number_patches, comm_landscape$forest_cover) 
# cor = -0.22, not correlated
```

## Fitting the models
```{r}
#| code-overflow: wrap
# Is species richness influenced by the number of patches in the landscape (Richness vs Number of patches)? 
mod1 <- lm(richness~number_patches, data = landscapes) 
summary(mod1) # r2-adj= 0.20, slope = -0.21, p = 0.03

# Generalized linear regression with poisson (count) distribution
mod2 <- glm(richness~number_patches, data = landscapes, family = "poisson") 
summary(mod2) # slope = -0.01, p = 0.03

AIC(mod1, mod2) # both models are equally plausible. I will keep mod1.

shapiro.test(mod1$residuals) # normal residuals (p=0.90)
bptest(mod1) # homoscedasticity (p=0.255)
```

## Plotting the graph
```{r, output=FALSE}
#| code-overflow: wrap
# Plotting the graph
landscapes$cor <- NA
landscapes$cor[1:8] <- "#4575b4"
landscapes$cor[9:12] <- "#fee090"
landscapes$cor[13:20] <- "#d73027"

plot2 <-
  ggplot(data = landscapes,
         mapping = aes(x = number_patches, y = richness)) +
  
  geom_smooth(method = "lm", fill = "#cccccc", alpha = 0.5,
              color = "#252525") +
  
  labs(x = "Number of islands in the landscape",
       y = "Number of ant species") +
  
  geom_image(
    data = tibble(number_patches = 55, richness = 51.9),
    aes(image = "ant_1.png"), size = 0.08 
  ) + # https://www.phylopic.org/images/3ae7cb42-8904-4775-87d6-968255913711/camponotus-vagus
  
  geom_point(size = 4, shape = 21,
             color = "#252525", fill = landscapes$cor,
             alpha = 0.7) +
  
  theme_bw(base_size = 10) +
  
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "#252525"),
        axis.title = element_text(colour = "#252525", face = "bold", size = 14),
        axis.title.x = element_text(vjust = 0, size = 14),
        axis.title.y = element_text(vjust = 2, size = 14),
        axis.text = element_text(colour = "#252525"),
        axis.ticks = element_line(colour = "#252525", size = 0.25),
        plot.margin = margin(t=10, r=10, b=100, l=10)) +
  
   coord_cartesian(clip = "off")


img10 <- rasterGrob(readPNG("graphical_abstract2.png"), interpolate = TRUE)
img30 <- rasterGrob(readPNG("landscapes_x-axis_3.png"), interpolate = TRUE)
img50 <- rasterGrob(readPNG("graphical_abstract2_3.png"), interpolate = TRUE)

rect1 <- rectGrob(gp = gpar(col = "#4575b4", fill = NA, lwd = 2))
rect2 <- rectGrob(gp = gpar(col = "#fee090", fill = NA, lwd = 2))
rect3 <- rectGrob(gp = gpar(col = "#d73027", fill = NA, lwd = 2))


plot3 <-
  plot2 +
  scale_x_continuous(limits = c(8, 58),
    breaks = c(10, 20, 30, 40, 50, 60),
    expand = expansion(mult = c(0, 0.05))) +
  
  scale_y_continuous(limits = c(25, 54),
    breaks = c(30, 35, 40, 45, 50),
    expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(clip = "off") +
  
  annotation_custom(img10, xmin = 9, xmax = 19, ymin = 14, ymax = 20) +
  annotation_custom(img30, xmin = 29, xmax = 39, ymin = 14, ymax = 20) +
  annotation_custom(img50, xmin = 49, xmax = 59, ymin = 14, ymax = 20) +
  
  annotation_custom(rect1, xmin = 9.8, xmax = 18.2, ymin = 14, ymax = 20) +
  annotation_custom(rect2, xmin = 29.8, xmax = 38.2, ymin = 14, ymax = 20) +
  annotation_custom(rect3, xmin = 49.8, xmax = 58.2, ymin = 14, ymax = 20) +
  
  annotation_custom(grid::linesGrob(arrow = arrow(length = unit(0.3, "cm")),
    gp = gpar(col = "#252525")),
  xmin = 10, xmax = 58, ymin = 21, ymax = 21) +
  
  annotation_custom(grid::textGrob(label = "Low\nfragmentation", x = unit(14, "native"), 
                                   y = unit(13, "native"),
                    gp = gpar(fontsize = 10, col = "#252525")),
                    xmin = 14, xmax = 14, ymin = 12, ymax = 12) +
  
  annotation_custom(grid::textGrob(label = "Moderate\nfragmentation", x = unit(34, "native"), 
                                   y = unit(13, "native"),
                                   gp = gpar(fontsize = 10, col = "#252525")),
                    xmin = 34, xmax = 34, ymin = 12, ymax = 12) +
  
  annotation_custom(grid::textGrob(label = "High\nfragmentation", x = unit(54, "native"), 
                                   y = unit(13, "native"),
                                   gp = gpar(fontsize = 10, col = "#252525")),
                    xmin = 54, xmax = 54, ymin = 12, ymax = 12)
```

```{r, fig.width=5.04, fig.height=5.35}
#| code-overflow: wrap
plot3

ggsave(
  plot = plot3,
  filename = "Richness_NumberPatches_v10.png", dpi = 500,
  width = 8 * 1.6, height = 8 * 1.7, units = 'cm')   

```

## Identifying the best model

To check which variable best explains the number of species in the landscapes, we used model selection analysis to identify the best-performing model. The best model included only number of patches.

```{r}
mod_global <- lm(richness ~ number_patches + forest_cover, data = landscapes, na.action = na.fail)
  
models <- dredge(mod_global)

models
```
