# The across-habitat heterogeneity

To ensure that the assumption of low and similar habitat heterogeneity was met we measured forest heterogeneity of landscapes using the standard deviation of Normalized Difference Vegetation Index (NDVI) across the 10 sampled points within each landscape. 
We also calculated the standard deviation of Enhanced Vegetation Index (EVI), as this index is recommended for tropical regions because NDVI tends to saturate in areas of high biomass [(Huete et al. 2002)](https://www.sciencedirect.com/science/article/pii/S0034425702000962?casa_token=bzwRRnSEmQUAAAAA:ZL-08r7mpEXPbNoyk1OkrxvogtazjvjdRB6_7l9thEOPUtvh1HcXIyUGIIMpX17HcDXJCIuVOQ).

I used Landsat Collection 9 Level 2 (30 m) images retrieved from [USGS](https://www.usgs.gov/centers/eros/science/usgs-eros-archive-landsat-archives-landsat-8-9-olitirs-collection-2-level-2). I followed the instructions on the same site to calculate the two vegetation index: [Normalized Difference Vegetation Index (NDVI)](https://www.usgs.gov/landsat-missions/landsat-normalized-difference-vegetation-index)(NDVI = (Band 5 - Band 4) / (Band 5 + Band 4)), and the [Enhanced Vegetation Index](https://www.usgs.gov/landsat-missions/landsat-enhanced-vegetation-index) (EVI = 2.5 * ((Band 5 – Band 4) / (Band 5 + 6 * Band 4 – 7.5 * Band 2 + 1))).


## Loading packages and importing data
```{r, output = FALSE}
#| code-overflow: wrap
library(terra)
library(sf)
library(raster)
library(dplyr)
library(ggplot2)
library(stats)
library(landscapemetrics)
library(LSRS)

# Importing data
rm(list = ls()); gc()
set.seed(13)

new_crs <- CRS("+proj=utm +south +zone=21 +datum=WGS84 +units=m +no_defs")

# Tile 1 (80% of the lake)
Balbina_red_1 <- rast("LC09_L2SP_231061_20240826_20240827_02_T1_SR_B4.TIF")
Balbina_red_1 <- project(Balbina_red_1, new_crs)
Balbina_nir_1 <- rast("LC09_L2SP_231061_20240826_20240827_02_T1_SR_B5.TIF")
Balbina_nir_1 <- project(Balbina_nir_1, new_crs)
Balbina_band2_1 <- rast("LC09_L2SP_231061_20240826_20240827_02_T1_SR_B2.TIF")
Balbina_band2_1 <- project(Balbina_band2_1, new_crs)

# Tile 2 (20% of the lake)
Balbina_red_4 <- rast("LC09_L2SP_230061_20240819_20240820_02_T1_SR_B4.TIF") 
Balbina_red_4 <- project(Balbina_red_4, new_crs)
Balbina_nir_4 <- rast("LC09_L2SP_230061_20240819_20240820_02_T1_SR_B5.TIF") 
Balbina_nir_4 <- project(Balbina_nir_4, new_crs)
Balbina_band2_4 <- rast("LC09_L2SP_230061_20240819_20240820_02_T1_SR_B2.TIF")
Balbina_band2_4 <- project(Balbina_band2_4, new_crs)

# Calculating NDVI
Balbina_ndvi_1 <- (Balbina_nir_1 - Balbina_red_1) / (Balbina_nir_1 + Balbina_red_1)
Balbina_ndvi_4 <- (Balbina_nir_4 - Balbina_red_4) / (Balbina_nir_4 + Balbina_red_4)
Balbina_ndvi_4 <- resample(Balbina_ndvi_4, Balbina_ndvi_1)
res(Balbina_ndvi_4) == res(Balbina_ndvi_1)
Balbina_ndvi <- terra::merge(Balbina_ndvi_1, Balbina_ndvi_4)
remove(Balbina_ndvi_1)
remove(Balbina_ndvi_4)
```

## Calculating NDVI for the sampled islands in the landscapes
```{r}
#| code-overflow: wrap
points <- read.csv("ants_env.csv")
points <- st_as_sf(points, coords = c("long_utm", "lat_utm"), crs = new_crs)

# NDVI per point
NDVI_1 <- extract(Balbina_ndvi, points, ID = FALSE)
NDVI_1 <- cbind(point_id = points$point_ID, NDVI_1)
colnames(NDVI_1)[2] = "NDVI"
NDVI_1 <- NDVI_1[-c(201:240),]
NDVI_1$landscape = points[-c(201:240),]$landscape

NDVI_1_sd <- tapply(NDVI_1$NDVI, NDVI_1$landscape, sd, na.rm = TRUE)
NDVI_1_sd <- as.data.frame(NDVI_1_sd)
colnames(NDVI_1_sd) <- "sd_NDVI"
NDVI_1_sd$landscape_id <- row.names(NDVI_1_sd)
NDVI_1_sd$landscape_id <- as.factor(NDVI_1_sd$landscape_id)

NDVI_1_sd$type <- NA
NDVI_1_sd$type[1:8] <- "FL"
NDVI_1_sd$type[9:12] <- "MS"
NDVI_1_sd$type[13:20] <- "SS"
NDVI_1_sd$type <- as.factor(NDVI_1_sd$type)

mod2 <- aov(NDVI_1_sd$sd_NDVI ~ NDVI_1_sd$type)
summary(mod2) # F=0.575, p=0.573
```

## Plotting the graph: NDVI
```{r}
#| code-overflow: wrap
plot2 <-
  ggplot(data = NDVI_1_sd,
         aes(x= type, y=sd_NDVI)) +
  
  labs(x = "Landscape type",
       y = "Standard deviation of NDVI") +
  
  geom_point(alpha = 0.4, position = position_jitter(width = 0.2)) + 
  
  stat_summary(fun = mean, geom = "point", color = "red", size = 2) + 
  
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", color = "red", width = 0.2) +
  
  scale_x_discrete(labels = c("Low \nfragmentation", "Moderate \nfragmentation", "High \nfragmentation")) +
  
  theme_bw(base_size = 10) +
  
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "black"),
        axis.title = element_text(colour = "black", face = "bold"),
        axis.text = element_text(colour = "black"),
        axis.ticks = element_line(colour = "black", size = 0.25),
        legend.position = "none") +
  
  annotate("text", x = 0.5, y = 0.025,
           hjust = 0, vjust = 0, size = 2,
           parse = T,
           label = as.character(expression(italic(F)*"-value = 0.57"))) +
  
  annotate("text", x = 0.5, y = 0.025,
           hjust = 0, vjust = 2, size = 2,
           parse = T, label = as.character(expression(italic(p)*"-value = 0.57")))
plot2

```

## Calculating EVI for the sampled islands in the landscapes
```{r, output=FALSE}
#| code-overflow: wrap
# LSRS uses the same formula suggested by USGS: 2.5*(NIR - Red) / (NIR + 6*Red - 7.5*Blue + 1)

Balbina_evi_1 <- LSRS::EVI(a = Balbina_nir_1, b = Balbina_red_1, c = Balbina_band2_1, Pixel.Depth = 32)

Balbina_evi_4 <- LSRS::EVI(a = Balbina_nir_4, b = Balbina_red_4, c = Balbina_band2_4, Pixel.Depth = 32)

Balbina_evi_4 <- resample(Balbina_evi_4, Balbina_evi_1)
res(Balbina_evi_4) == res(Balbina_evi_1)
Balbina_evi <- terra::merge(Balbina_evi_1, Balbina_evi_4)
plot(Balbina_evi)


# EVI per point
points <- read.csv("ants_env.csv")
points <- st_as_sf(points, coords = c("long_utm", "lat_utm"), crs = new_crs)

EVI_1 <- extract(Balbina_evi, points, ID = FALSE)
EVI_1 <- cbind(point_id = points$point_ID, EVI_1)
colnames(EVI_1)[2] = "EVI"
EVI_1$landscape = points$landscape

EVI_1_sd <- tapply(EVI_1$EVI, EVI_1$landscape, sd, na.rm = TRUE)
EVI_1_sd <- as.data.frame(EVI_1_sd)
colnames(EVI_1_sd) <- "sd_EVI"
EVI_1_sd$landscape_id <- row.names(EVI_1_sd)
EVI_1_sd$landscape_id <- as.factor(EVI_1_sd$landscape_id)

EVI_1_sd$type <- NA
EVI_1_sd$type[1:8] <- "FL"
EVI_1_sd$type[9:12] <- "MS"
EVI_1_sd$type[13:20] <- "SS"
EVI_1_sd$type <- as.factor(EVI_1_sd$type)
```

```{r}
#| code-overflow: wrap
mod3 <- aov(EVI_1_sd$sd_EVI ~ EVI_1_sd$type)
summary(mod3) #F=0.415, p=0.67
```

## Ploting the graph: EVI
```{r}
#| code-overflow: wrap
plot2 <-
  ggplot(data = EVI_1_sd,
         aes(x= type, y=sd_EVI)) +
  
  labs(x = "Landscape type",
       y = "Standard deviation of EVI") +
  
  geom_point(alpha = 0.4, position = position_jitter(width = 0.2)) + 
  
  stat_summary(fun = mean, geom = "point", color = "red", size = 2) + 
  
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
               geom = "errorbar", color = "red", width = 0.2) +
  
  scale_x_discrete(labels = c("Low \nfragmentation", "Moderate \nfragmentation", "High \nfragmentation")) +
  
  theme_bw(base_size = 10) +
  
  theme(panel.grid = element_blank(),
        panel.border = element_rect(colour = "black"),
        axis.title = element_text(colour = "black", face = "bold"),
        axis.text = element_text(colour = "black"),
        axis.ticks = element_line(colour = "black", size = 0.25),
        legend.position = "none") +
  annotate("text", x = 0.5, y = 0.018,
           hjust = 0, vjust = 0, size = 2,
           parse = T,
           label = as.character(expression(italic(F)*"-value = 0.41"))) +
  
  annotate("text", x = 0.5, y = 0.018,
           hjust = 0, vjust = 2, size = 2,
           parse = T, label = as.character(expression(italic(p)*"-value = 0.67")))
plot2
```
